---
title: "Difference-in-differences Designs"
date: "`r Sys.Date()`"
link-citations: yes
bibliography: ../DirectEffects.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Difference-in-differences Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Difference-in-differences Designs

When researchers have pretreatment measurements of the outcome and the mediator, they can use the difference-in-differences (DID) approach to estimating controlled direct effects of @BlaGlyHil24. This approach can take full advantage of doubly robust estimation as with other models. 

Generally speaking, the DID approach and the selection-on-observable approach are very similar, with a few differences. First, the outcome should represent a difference between the pretreatment and posttreatment measurements of the outcome. Second, one must specify the name of the variable measuring the pretreatment values of the mediator since the estimates will be conditional on this variable. Third, there are two different quantities of interest in this setting that can be estimated, depending on what one wants to condition on and what assumptions one wants to make. 


In this article, we demonstrate the estimation of CDEs in the DID setting using data from the field experiment of @BroKal16. This experiment consisted of one baseline survey (`t0`) and four posttreatment surveys (`t1` through `t4`). In the experiment, the authors randomly assigned an door-to-door intervention to either engage in a conversation promoting transgender rights (`treated == 1`) or a control conversation about recycling (`treated == 0`). Our goal is to estimate the direct effect of this treatment fixing values of posttreatment subjective feelings about transgender people (`therm_trans_t2`, measured in wave 2). The outcome of interest is support for a transgender nondiscrimination law (`nondiscrim_law_t3`, measured in wave 3). Baseline covariates consist of several survey measures (labeled with `t0`) and several variables from the matched voter file (labeled `vf`). These baseline measures include pretreatment measurements of the mediator (`therm_trans_t0`) and the outcome (`nondiscrim_law_t0`). For purposes of estimation, we will use the difference between the pre and posttreatment outcomes (`nondiscrim_law_diff`) as the outcome in the models below.  We measure intermediate confounders at from wave 1 (`t1`). 

We can see a brief view of the data below:

```{r loadpkg, echo = FALSE, include = FALSE}
library(tidyverse)
library(DirectEffects)
data(transphobia)
```


```{r}
transphobia
```

## Specifying conditioning sets 

One key step in estimating CDEs is to specify the covariates that we will condition on for each propensity score and outcome regression. We can specify in the relevant call to `treat_model` or `outreg_model`, but it is often useful to define the formulas separately (perhaps to use them in several different models). 


```{r}
## Propensity score model for treatment
ps_t_mod <- treated ~ therm_trans_t0

## Propensity score model for treatment
ps_m_mod <- therm_trans_t2 ~ treated + therm_trans_t0 + gender_norm_moral_t0 +
  therm_obama_t0 + vf_democrat + vf_female + vf_black + vf_hispanic +
  gender_norm_moral_t1 + therm_obama_t1 

## Outcome models for treatment (short regression)
or_t_mod <- nondiscrim_law_diff ~  treated + therm_trans_t0 +
  gender_norm_moral_t0 + therm_obama_t0 + vf_democrat + vf_female +
  vf_black + vf_hispanic


## outcome model for mediator (long regression)
or_m_mod <- nondiscrim_law_diff ~ treated + therm_trans_t0 + therm_trans_t2 +
  gender_norm_moral_t0 + therm_obama_t0 + vf_democrat + vf_female + vf_black +
  vf_hispanic  + gender_norm_moral_t1 + therm_obama_t1
```

Here, the propensity score model for treatment only includes the baseline mediator as treatment is randomly assigned. For the outcome regression for treatment, we do include baseline covariates for efficiency reasons. For the mediator models, we include baseline and intermediate covariates. 



## Estimating the ACDE

We now turn to estimation. The main building block for the DID approach is the `cde_did_aipw()` function. Like the other `cde_aipw()` function, there is the `trim` argument that trims the propensity score weights to keep estimates more stable and the `aipw_blip` which is a logical that indicates whether the estimator should use a doubly-robust estimator for the intermediate "blipped-down" regression functions.  

There are several arguments unique to the DID setting. First, the `base_mediator` argument specifies what variable in the data frame will contain the baseline (that is, pretreatment) mediator. The estimates will be conditional on this variable so it should be discrete. Second, the `on_treated` logical argument indicates if the estimand should further be conditional on being treated and having a constant value of the mediator over time.

Below we specify the entire chain of estimators, using `multinom` for the mediator because it has three levels (feeling cool, neutral, and warm toward transgender people). 

```{r}
out_dr_ml <- cde_did_aipw(base_mediator = therm_trans_t0, trim = c(0.01, 0.99)) |>
  set_treatment(treated) |>
  treat_model(engine = "logit", ps_t_mod) |>
  outreg_model(engine = "lm", or_t_mod, separate = FALSE) |>
  set_treatment(therm_trans_t2) |>
  treat_model(engine = "multinom", ps_m_mod, separate = FALSE, include_past = FALSE) |>
  outreg_model(engine = "lm", or_m_mod, separate = FALSE, include_past = FALSE) |>
  estimate(nondiscrim_law_diff ~ treated, data = transphobia, n_folds = 5)
```

Notice that we have set `include_past` to `FALSE` for the mediator models because we have explicitly included those covariates in our model specifications. In addition, we use the difference between posttreatment and pretreatment outcomes as the depedent variable in this specification. 


Using the default `on_treated == FALSE` implies that we are estimating what @BlaGlyHil24 calls the *baseline conditional* ACDE or ACDE-BC. We can view the output from these, with all of the ACDE-BCs using `summary()`. This also displays the marginalized ACDE-BC that averages the conditional effects over the distribution of the baseline mediator. 

```{r}
summary(out_dr_ml)
```


Here we can see that there is a statistically significant ACDE-BC for the neutral group (m = 1), but not for the cool or warm groups.  One can also use the `tidy()` function from the **broom** package to easily extract information:

```{r}
tidy(out_dr_ml)
```


This allows for easy plotting of the results using **ggpplot2**:

```{r fig.alt = "Estimated ACDEs from a DID analysis"}
tidy(out_dr_ml) |>
  ggplot(aes(x = term, y = estimate)) +
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_minimal()
```

Setting `on_treated` to `TRUE`, on the other hand, estimates the path-conditional ACDE or ACDE-PC.  

```{r}
out_dr_ml_pc <- cde_did_aipw(base_mediator = therm_trans_t0, on_treated = TRUE,
                             trim = c(0.01, 0.99)) |>
  set_treatment(treated) |>
  treat_model(engine = "logit", ps_t_mod) |>
  outreg_model(engine = "lm", or_t_mod, separate = FALSE) |>
  set_treatment(therm_trans_t2) |>
  treat_model(engine = "multinom", ps_m_mod, separate = FALSE, include_past = FALSE) |>
  outreg_model(engine = "lm", or_m_mod, separate = FALSE, include_past = FALSE) |>
  estimate(nondiscrim_law_diff ~ treated, data = transphobia, n_folds = 5)

summary(out_dr_ml_pc)
```



## References
